---
title: "DENVar: A global Dynamic Estimate of NDVI Variance"
subtitle: Appendix A
author:
  - name: Stefano Mezzini
    institute: [braes, biol, pois]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - pois: Poisson Consulting Ltd., Nelson, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'ndvi-stochasticity.bib'
csl: 'remote-sensing-of-environment.csl'
suppress-bibliography: false # if refs can only be in main text
urlcolor: cyan
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A.\!\!} # for "Fig. A.x"
    - \usepackage{float} # for fig.pos = '!H'
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: false # for submitting TeX file
editor_options:
  chunk_output_type: console
---

\clearpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center', fig.pos = '!h')
```

```{r packages, warning=FALSE}
library('knitr')
```

\noindent Fig. A.1 below illustrates the color palettes used in the main manuscript and this appendix, for each type of vision (trichromatic, deuteranope, protanope, tritanope, and achomatic). The NDVI palette is a modified version of the Bukavu palette by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018].

The palette for Fig. A.4 is a modified version of a pallette by Masataka Okabe and Kei Ito [@okabe_color_2008]. All other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018]. All original palettes are available via the `khroma` package [@frerebeau_khroma_2024] for `R` [@r_core_team_r_2024]. See the Code Availability statement in the main text for the code used to produce the figures. The shapefile of WWF ecozones and biomes was obtained from the work of @olson_terrestrial_2001.

\singlespacing

```{r color-palettes, fig.cap="Color palettes used in the main manuscript and the figures below, along with simulated color-deficient views."}
include_graphics('../figures/input-data/color-palettes.pdf')
```

<!-- \clearpage <!-- for cleaner pagination and to keep groups and biomes on the same page-->

```{r prop-water, fig.cap="Robinson projection of the proportion of water in each pixel. Black lines indicate WWF ecozones."}
include_graphics('../figures/input-data/proportion-water.pdf')
```

<!-- \clearpage -->

```{r ndvi-ecdf, fig.cap="Monthly empirical cumulative density fuctions of NDVI across latitude. The white lines indicate the 99$^{\\text{th}}$ quantiles of NDVI at each latitude (thinner lines) and floored latitude (thicker lines). The red areas indicate the NDVI values $> 0$ removed in each month. No values at latitudes $< 60^\\circ$S were removed since Antarctica was not included in the analyses."}
include_graphics('../figures/global-diagnostics/monthly-latitudinal-ecdf.pdf')
```

<!-- \clearpage -->

```{r data-groups, fig.cap="Robinson projection of the groups used to avoid data size limitations in $\\texttt{R}$. Black lines indicate WWF ecozones. The Antarctic group was excluded from the analyses.", out.width='95%'}
include_graphics('../figures/input-data/data-groups.pdf')
```

<!-- \clearpage -->

```{r biomes, fig.cap="Robinson projection of WWF biomes. Black lines indicate WWF ecozones."}
include_graphics('../figures/input-data/biomes.pdf')
```

<!-- \clearpage -->

```{r n-rasters, fig.cap="Number of days with a raster within each 4-day period used for temporal aggregation of the data."}
include_graphics('../figures/input-data/n-rasters-time.pdf')
```

<!-- \clearpage -->

```{r elevation, fig.cap="Robinson projection of elevation above sea level. All elevations below the lowest altitude in the Dead Sea (-432 m) were set to -432 m to remove excessively negative altitudes near coast lines. The $\\texttt{elevatr}$ package for $\\texttt{R}$ did not provide elevations for the southernmost region of Antarctica. Black lines indicate WWF ecozones."}
include_graphics('../figures/input-data/elev-m.pdf')
```

\clearpage <!-- refs on separate page -->

# References {-}

