---
title: \large **ACRONYM**: A new global measure of environmental unpredictability 
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'ndvi-stochasticity.bib'
csl: 'freshwater-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: false # for submitting TeX file
editor_options:
  chunk_output_type: console
---

```{r, echo=FALSE}
abstract <- ""
```

<!-- elements of the title page -->

\clearpage

\noindent \textbf{Article type}: Research article

\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`

\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`

\noindent \textbf{Figures}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{Tables}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{References}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{Appendices}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{Key words:} remote sensing, NDVI, environmental stochasticity, \textcolor{red}{\textbf{XXX}}

<!-- start main body on a new page -->

\newpage

\doublespacing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
```

\linenumbers

# Abstract

\noindent `r abstract`

\newpage

# Acronym:

* NVGI: NDVI Variance GAM-smoothed Index
* GAM-Estimated 
* Ginli: GAM-Inferred NDVI ??? Index
* env var

potential words starting with a vowel: Index, Estimate(d), unpredictability

potential words starting with a consonant: GAM, smooth(ed), NDVI, stochasticity, variance, 

Global GAM 



# Introduction

https://www.sciencedirect.com/science/article/abs/pii/S0034425718305625
https://silvis.forest.wisc.edu/data/dhis/

# Methods

maps of input data

NDVI palette is colorblid-friendly

# Results

output of `draw()`

hex plots of mean and var vs common metrics:
* temperature
* seasonal temperature range
* precip
* seasonal temperature range
* spp diversity
* net primary productivity
* hfi
* max animal weight

# Discussion

* stable states
* green wave surfing
* migration routes are key habitat (Ortega et al., in prep), especially in stochastic environments
* @sawyer_migratory_2019
* https://www.sciencedirect.com/science/article/pii/S0960982220308484
* https://onlinelibrary.wiley.com/doi/full/10.1111/gcb.15169
* industrial development affects animals' ability to respond to both predictable and unpredictable change (https://www.nature.com/articles/s41559-022-01887-9)
* effects of sudden deep snow in red desert on proghorn (https://www.sciencedirect.com/science/article/abs/pii/S0960982225002957)
* https://www.nature.com/articles/s41467-023-37750-z
* https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.4238

* smoothness of mu impacts smoothness and size of s2: depends on animals' ability to respond to, learn, and predict E(R)

# Useful references

* `{khroma}`: @frerebeau_khroma_2024
* @keith_predicting_2008
* @chevin_adaptation_2010
* @rickbeil_plasticity_2019
* @mueller_search_2008
* @pettorelli_using_2005
* @keys_machine-learning_2021
* @nilsen_can_2005
* @stephens_optimal_1982
* @merkle_large_2016
* @tian_evaluating_2015
* @huang_commentary_2021
* @fan_global_2016
* @wang_stochastic_2019
* @pease_ecological_2024
* @xu_plasticity_2021
* @mezzini_how_2025

# References
